Epistemic Semantics in Guarded String Models\thanks{Thanks guys.}

### 2021-01-14 tex/pdraft.tex

\author{Eric Campbell \\
  Cornell University\\
  \texttt{ehc86@cornell.edu} \\\And
  Mats Rooth \\
  Cornell University\\
  \texttt{mr249@cornell.edu} \\}

@@

\begin{document}
\maketitle
\begin{abstract}
  Constructive and computable multi-agent epistemic possible worlds
  models are defined, where possible worlds models are guarded string
  models in an epistemic extension of Kleene Algebra with Tests.  The
  account is framed as a formal language Epik (Epistemic KAT) for
  defining such models.  The language is implemented by translation
  into the finite state calculus, and alternatively by modeling
  propositions as lazy lists in Haskell. The syntax-semantics
  interface for a fragment of English is defined by a categorial
  grammar.
\end{abstract}

@@

The approach taken here synthesizes five antecedents in a certain way.
John McCarthy's {\em Situation Calculus} is the source of the idea of
constructing possible worlds as event sequences
\citep{mccarthy1963situations,reiter2001knowledge}.  The algebraic
theory of {\em Kleene Algebra with Tests} characterizes algebras with
elements corresponding to propositions and event types in our
application \citep{kozen2001automata}.  {\em Action models} in dynamic
epistemic semantics introduce the technique of constructing epistemic
models from primitive alternative relations on events, in order to
capture the epistemic consequences of perceptual and communicative
events \citep{baltag1999logic}.  Literature on {\em finite state
  methods in linguistic semantics} has used event strings and sets of
event strings to theorize about tense and aspect in natural language
semantics
\citep{fernando2004finite,fernando2007observing,carlson2009tense} and
to express intensions \cite{fernando2017intensions}.  Work on {\em
  finite state intensiononal semantics} has investigated how to do the
semantics of intensional complementation in a setting where
compositional semantics is expressed in a finite state calculus
\citep{rooth2017finite,collard2018finite}.

A running example of an event-sequence model is {\em The Concealed
  Coin}.  Amy and Bob are seated at a table.  There is a coin on the
table under a cup, heads up.  The coin could be heads-up $H$ or
tails-up $T$, and neither agent knows which it is.  This initial
situation is possible world $w_1$.  Two additional worlds $w_2$ and
$w_3$ are defined by sequencing events after the initial state, with
events interpreted as in \ex{coin1}. The truth values for English
sentences shown in \ex{coin3} are observed, where $0$ stands for
falsity and $1$ for truth.

@@

\begin{example} \label{coin2}
  \(w_2 = w_1\apH \hfill
  w_3 = w_2\bpH \hfill
  w_4 = w_3\atHT\btTH\bpH\)
\end{example}

@@

A sequence such as $\bar{h}{t}$ can be viewed both as a formula and as
a valuation of primitive propositions, which we use to describe world
state. The primitives are listed in fixed order, and left unmarked
(indicating true) or marked with the overbar (indicating false).
Since a coin is heads or tails but not both, we want to allow the
valuations $h\bar{t}$ and $\bar{h}t$, and disallow $ht$ and
$\bar{h}\bar{t}$.\footnote{Two Booleans constrained by a state formula, rather than one Boolean, are stipulated here to
 illustrate state formulas.
}  This is enforced by a {\em state formula}, which is
a Boolean formula, in this case the one given on the second line of
\ex{effect}.  Where {\sf{B}} is a set of state primitives and $\phi$ is a
 constraint over {\sf{B}},
${\cal{A}}_{\mbox{\sf\tiny{B}}}^{\phi}$ is the set of valuations of
{\sf{B}} that make formula $\phi$ true. Valuations are called atoms,
because they correspond to the atoms of a Boolean algebra of tests
\citep{kozen2001automata}.

@@

Epik is a specification language for possible worlds models that
includes declarations of events and states, state formulas, effect
formulas, and additional information.  Figure \ref{epikfigure} shows
an Epik program that describes a possible worlds model for two agents
with information about one coin, events of the agents semi-privately
looking at the coin, and events of secretly turning the coin.  The
line beginning with {\tt state} enumerates ${\sf{B}}$. The line
beginning with {\tt restrict} gives the state formula.  The lines
beginning with {\tt event} declare events and their effect formulas.
Finally the lines beginning with {\tt agent} define {\em event
  alternative} relations for agents.  Each clause with an arrow has a
single event symbol on the left, and a disjunction of alternative
events on the right of the arrow. The interpretation of Amy's
alternatives for $b_1$ (Bob peeks at heads), is that when $b_1$
happens, for Amy either $b_1$ or $b_0$ (Bob peeks at tails) could be
happening. Her alternatives for $a_{01}$ and $a_{10}$ (she turns the
coin over in one state or the other) are $a_{10}$ and $a_{01}$,
indicating that she doesn't know, \emph{a priori}, whether she's
turning the coin from $H$ to $T$ or from $T$ to $H$. When Bob secretly
turns the coin over, in event $b_{10}$ or $b_{01}$, she doesn't know
anything has happened, so her alternative is the ``no-operation'' or
``no-information'' event event $o$. Bob's event relation is symmetric.

@@

This paper focuses on defining a concrete possible worlds model from an Epik
specification.  The models are an extension of guarded-string models for Kleene
Algebra with Tests (KAT). This is an algebraic theory that has model classes
including guarded string models, relational models, finite models, and matrix models.
Our definitions and notation follow \cite{kozen2001automata}.  We add syntax and
semantics to cover multi-agent epistemic semantics.

Guarded strings over a finite alphabet ${\sf{P}}$ are like ordinary strings, but with
atoms over a set ${\sf{B}}$ alternating with the symbols from ${\sf{P}}$.  In the
algebra described by Figure \ref{epikfigure}, ${\sf{P}}$ is the set of events
$\{a_1,a_0,b_1,b_0,a_{10},a_{01},b_{10},b_{01},o\}$, and ${\sf{B}}$ is $\{h,t\}$.  As
we already saw in \ex{effect}, $\mathcal{A}^\varphi_{\mathsf B}$ is
$\{h\bar{t},\bar{h}t\}$, for which we use the shorthand $\{H,T\}$.  A guarded string
over ${\sf{P}}$ and ${\sf{B}}$ is a string of events from ${\sf{P}}$, alternating
with atoms over ${\sf{B}}$, and beginning and ending with atoms.  In this
construction, $w_1=H$, $w_2=Ha_1H$, $w_3=Ha_1Hb_1H$, and
$w_4=Ha_1Hb_1Ha_{10}Tb_{01}Hb_1H$.

@@

Rather than individual guarded strings, elements of a guarded string
model for KAT are sets of guarded strings.  In our application, these
elements have the interpretation of propositions, which are sets of possible
worlds.  In a free guarded string model for KAT, any event can be
adjacent to any atom in a guarded string that is an element of the
underlying set for the algebra.  We instead impose the constraints
coming from the state and effect formulas.  \ex{wf1} defines the
well-formed guarded strings determined by an Epik specification.  Condition (i)
says that each atom is consistent with the state constraint, and condition (ii)
says that each constituent token event $\alpha_ie_i\alpha_{i+1}$ is
consistent with the effect constraint on $e_i$.\footnote{
An alternative is to define equations such as
$\bar{\phi}=0$ (from the state formula $\phi$)
and $a_1\,{=}\,ha_{1}h$ (from the effect formula $h\mathbin{:}h$ for event $a_1$),
and construct a quotient algebra from the equivalence relation generated
by these equations.  This results in equating sets of guarded strings in the free algebra that
differ by guarded strings that are ill-formed according to the state and
effect formulas.  In the development in the text, we instead use
a set of guarded strings that are well-formed according to the state
and effect formulas as the representative of the equivalence class.
}

@@

The construction so far defines a set of worlds from an Epik specification.  Normally
the set is countably infinite, though some choices of effect formulas can result in a
finite set of worlds.  The next step is to define an alternative relation $R_a$ on
worlds for each agent $a$. This will result in a general modal frame
$\tuple{W,R_1,...,R_n,K}$ consisting of a set of worlds, a world-alternative relation
for each agent, and a set $K$ of propositions, where each proposition is a subset of
$W$ \citep{chagrov1997modal}.\footnote{As explained in the next section, $K$ will not
  be the power set of $W$, rather it will consists of the regular subsets of $W$.}
An Epik specification defines an alternative relation on bare events for each agent
$a$, which we notate as ${R}_{a}$.  This should be lifted to a relation $\hat{R}_a$
on worlds.  The basic idea is that when a world $w$ is incremented with an event $e$,
in the resulting world $w \diamond e$, epistemic alternatives for agent $a$ are of
the form $w' \diamond e'$, where $w'$ is an alternative to for $a$ in $w$, and $e'$
is and event-alternative to $e$ for $a$.\footnote{In this it is important that the
  event-alternative relation for an agent is constant across worlds. We anticipate
  that the definition given here produces results equivalent to what is found in
  literature on event alternatives in dynamic epistemic semantics, though we have not
  verified this.  That literature primarily focuses on mapping an epistemic model for
  a single time and situation to another, and uses general first-order models, rather
  than guarded string models. See \citet{baltag1999logic}, \citet{van2007dynamic},
  and articles in \citet{ditmarsch2015handbook}.  This literature is motivated by
  epistemic logic and AI planning, rather than computable possible worlds models in
  natural language semantics.  } This needs to be implemented in a way that takes
account of pre- and post-conditions for events. For this, our approach is to refer
the definition of well-formed guarded strings.  \ex{raise1} defines an epistemic
alternative relation on worlds from an alternative relation on bare events.

@@

Consider a scenario like the one from Figure 1, but with an additional agent Cal. The
base world $T b_0 T c_0 T$ is one where the coin is tails, and first Bob looks at
tails, and them Cal looks at tails. The first event $b_0$ has the alternatives $b_0$
and $b_1$ for Amy, and the second event $c_0$ has the alternatives $c_0$ and $c_1$
for Amy.  This results in four combinations $b_0c_0$, $b_0c_1$, $b_1c_0$, and
$b_1c_1$. But these are filtered by post- and pre-conditions of events in the
alternative world, so that the set of alernatives for Amy in $T b_0 T c_0 T$ is $\{T
b_0 T c_0 T,H b_1 H c_1 H\}$, with two world-alternatives instead of four.

\begin{figure}
  \fbox{
    \begin{minipage}{\columnwidth}
      \(\begin{array}{llll}
      \multicolumn{4}{l}{\text{events}~~e\in \mathsf{P}} \\
      p,q &\multicolumn{3}{l}{::= e\,|\,\sigma\,|\,p+q\,|\,pq\,|\,p^*\,|\,\neg p\,|\, \Diamond_ap} \\
      \Box_a p & \triangleq \neg\Diamond_a\neg p &
      \bullet & \triangleq \sum_{e \in \mathsf P}e \\      
      p \wedge q &  \triangleq \neg(\neg p + \neg q) &
      p \to q & \neg p + q      
      \end{array} \)
  \end{minipage}}

@@

\section{The logical language of Epistemic KAT}
The standard language for Kleene algebra with tests has the signature
$\tuple{K,+,\cdot,*,\bar{},0,1}$ \citep{kozen2001automata}.  In a guarded string
model for KAT, $K$ is a set of sets of guarded strings, $+$ is set union, the
operation $\cdot$ is fusion product raised to sets, $*$ is Kleene star, the operation
$\;\bar{}\;$ is complement for tests, $0$ is the empty set, and $1$ is the set of
atoms.\footnote{$0$ has the dual role the identity for $+$ (union), and as False for
  operations on tests.  $1$ has the dual role of the identity for product (fusion
  product raised to sets), and True for tests.  } To this we add a unary
modal operation $\Diamond_a$ for each agent, and a unary complement operation $\neg$
on elements of $K$.  Intuitively, $\Diamond_a p$ is the set of
worlds where proposition $p$ is epistemically possible for agent $a$.  Propositional
complement is included because natural languages have sentence negation.  In
addition, universal box modalities are defined as duals of existential diamond
modalities.

With modalities and propositional negation added, the signature of $n$-agent
epistemic KAT is $\tuple{K,+,\cdot,*,\bar{},0,1,\neg,\Diamond_1,
  ..., \Diamond_n}$. Figure \ref{termfigure} defines the syntax of the language.
Juxtaposition is used for product.  Terms in this language
are used to represent the proposositional semantic values of English sentences.
\ex{interpretation1} gives some examples.  To explain the first one, $\bullet$ as
defined in Figure \ref{fig:epik-grammar} is the disjunction of the primitive events.
Since a world is a well-formed sequence of events, $\bullet^*$ is the set of worlds.
Multiplying by the state symbol $h$ in the term $\bullet^* h$ has the effect of
conjoining $h$ with the atom at the end of the world.  So $\bullet^* h$ is the set
of worlds where the coin ends heads-up.

@@

{\em Standard Epistemic Modalities}
Using our existential modal primitive $\Diamond_a$, and the dual
encodings of $\Box_a$ and $\wedge$, we can encode the standard modal
operators expressing knowledge $(\K_a)$ and belief $(\B_a)$
as in \ex{encodings}.\footnote{Deeper analyisis of the lexical semantics
  of {\em know} requires adding modeling of presupposition
  \citep{collard2018finite}. The grammar fragment in Section 6 does not model
  the presupposition of {\em know}, except as
  an entailment.}

@@

Summing up, given an Epik specification $\epikspec$, term $p$ (as defined
syntactically in Figure \ref{termfigure}) is interpreted as a set of guarded strings
$\sem{p}^{\epikspec}$.  Let $K^{\epikspec}$ be the sets that are interpretations of
terms.  Then
$\tuple{K^{\epikspec},+,\cdot,*,\bar{},0,1,\neg,\Diamond_{a_1},...,\Diamond_{a_n}}$
is a concrete guarded string interpretation for the signature of epistemic KAT, with
operations as in Figure \ref{fig:interp2} (e.g. the binary operation $+$ is union,
and the unary operation $\Diamond_a$ is pre-image relative to $\hat{R}_a$).  This
provides a concrete $n$-agent general modal frame
$\tuple{\sem{\bullet^*}^{\epikspec}, \hat{R}_1,...,\hat{R}_n,K^{\epikspec}}$.  The
frame consists of a set of worlds, an epistemic-alternative relation for each agent,
and a set of propositions, each of which is a set of worlds.  It is used as a target
for natural-language interpretation in Section 6.

@@

\section{Translation into the finite state calculus}
The finite state calculus is an algebra of regular sets of strings and
regular relations between strings that was designed for use in
computationonal phonology and morphographemics
\citep{kaplan1994regular,beesley2003finite}.  Current implementations
allow for the definition of functions on regular sets and relations
\citep{hulden2009foma,linden2009hfst,karttunen2010update}.  Such definitions are used
here to construct of a model for epistemic KAT inside the finite state
calculus. We describe our translation from Epik logical terms to Fst terms
here.

@@

\begin{hangingpar}
\begin{verbatim}
define Cn(X,Y) 
  [[[X Y] & Wf0] .o. Squash].l;
\end{verbatim}
KAT product.
\end{hangingpar}

@@

\begin{hangingpar}
\begin{verbatim}
define Kst(X) St | Kpl(X);
\end{verbatim}
KAT Kleene plus and Kleene star. The Fst operation {\tt |} is union.
\end{hangingpar}
\end{tcolorbox}
\caption{\raggedright Definition in Fst of KAT product and star.
  %% Where {\tt X} and {\tt Y} are regular sets and
  %% {\tt R} and {\tt S} are regular relatiions,
  %% {\tt X{\&}Y} is the intersection of {\tt X} and {\tt Y},
  %% {\tt X|Y} is the union of {\tt X} and {\tt Y},
  %% {\tt \~{}X} is the complement of {\tt X},
  %% {\tt R{.o.}S} is the composition of {\tt R} and {\tt S},
  %% {\tt R{.}l} is the co-domain of {\tt R}, and
  %% {\tt {\$}X} is the set of strings that have a substring in $X$.   
}
\label{fstfigure1}
\end{figure}

@@

Translating the Epik terms $0$, $1$, $b$, and $e$ are
straightforward: we simply convert the atoms as previously described,
and decorate the events $e$ with their compatible atoms. For example $a_1$ becomes
an Fst term denoting
$\{01a_101\}$. Fst has built-in operations of union (\texttt{|}),
and intersection (\texttt{\&}), which are
the sum and intersection operations in the guarded string algebra as represented
in Fst.
Fst set difference (\texttt{-}) is used to define propositional complement as the
difference between the set of worlds and the argument.

Defining KAT product using Fst's set-lifted string concatenation (denoted by
juxtaposition $\texttt{X}~\texttt{Y}$) requires more care. Naively concatenating
strings with atoms (Boolean vectors) at both ends doubles atoms at the juncture, and
does not enforce the requisite atom equality.  To implement KAT product, we define
the binary operation \texttt{Cn}, which concatenates strings in the string algebra,
removes strings with non-matching atoms, and then deletes the second of two atoms to
create a set of well-formed guarded strings.  See Figure \ref{fstfigure1}. {\tt Wf0}
is the set of ordinary strings that does not contain unequal pairs of atoms, as
defined using Fst's containment operator \texttt{\$}. The \texttt{Squash} relation
uses Fst's rewrite notation to delete atoms (elements of \texttt{St}) that are
preceded by another atom. \footnote{ This is a non-equal length regular relation. The
  finite state calculus includes such relations, and they can be used with relation
  composition and relation domain and co-domain.  They are restricted in that the
  complement and set difference for non-equal length relations is not defined.
  Epistemic alterative relations are equal-length relations.} This relation is
applied via the relational composition (\texttt{.o.}) and codomain (\texttt{.l})
operators.
%% Deletion is accomplished by a
%% re-write rule in the finite state calculus, which is a notation for
%% defining regular relations by contextually constrained substitutions.
%% In this case, is {\tt Squash} is a regular relation that deletes an
%% atom (a sequence of 0's and 1's of a certain length) when it follows
%% an atom.

>>>>

\author{Eric Hayden Campbell \\
  Cornell University\\
  \texttt{ehc86@cornell.edu} \\\And
  Mats Rooth \\
  Cornell University\\
  \texttt{mr249@cornell.edu} \\}

@@

\begin{document}
\maketitle
\begin{abstract}
  Constructive and computable multi-agent epistemic possible worlds
  models are defined, where possible worlds models are guarded string
  models in an epistemic extension of Kleene Algebra with Tests.  The
  account is framed as a formal language Epik (Epistemic \textsc{Kat}) for
  defining such models.  The language is implemented by translation
  into the finite state calculus, and alternatively by modeling
  propositions as lazy lists in Haskell. The syntax-semantics
  interface for a fragment of English is defined by a categorial
  grammar.
\end{abstract}

@@

The approach taken here synthesizes five antecedents.
John McCarthy's {\em Situation Calculus} is the source of the idea of
constructing possible worlds as event sequences
\citep{mccarthy1963situations,reiter2001knowledge}.  The algebraic
theory of {\em Kleene Algebra with Tests} characterizes algebras with
elements corresponding to propositions and event types in our
application \citep{kozen2001automata}.  {\em Action models} in dynamic
epistemic semantics introduce the technique of constructing epistemic
models from primitive alternative relations on events, in order to
capture the epistemic consequences of perceptual and communicative
events \citep{baltag1999logic}.  Literature on {\em finite state
  methods in linguistic semantics} has used event strings and sets of
event strings to theorize about tense and aspect in natural language
semantics
\citep{fernando2004finite,fernando2007observing,carlson2009tense} and
to express intensions \cite{fernando2017intensions}.  Work on {\em
  finite state intensional semantics} has investigated how to do the
semantics of intensional complementation in a setting where
compositional semantics is expressed in a finite state calculus
\citep{rooth2017finite,collard2018finite}.

A running example of an event-sequence model is {\em The Concealed Coin}. Amy
and Bob are seated at a table. There is a coin on the table under a cup, heads
up. The coin could be heads-up $H$ or tails-up $T$, and neither agent knows
which it is. This initial situation is possible world $w_1$. We can construct
additional worlds by sequencing events like those shown in \ex{coin1}.

@@

The worlds $w_1,\ldots,w_4$ in \ex{coin2} are examples constructed from the
events in \ex{coin1}. Juxtaposition indicates the order and occurence of events,
for example $w_3$ indicates that after the events of $w_2$, event $b_1$ occurs,
fully, this indicates beginning in world $w_1$, Amy surreptitiously peeking at
a heads-up coin, and then Bob doing the same.

\begin{example} \label{coin2}
  \(w_2 \triangleq w_1\apH \hfill
  w_3 \triangleq w_2\bpH \hfill
  w_4 \triangleq w_3\atHT\btTH\bpH\)
\end{example}

The truth values for English sentences are observed at the end of the event
sequences specified by $w_1,\ldots, w_4$, where $0$ stands for falsity and $1$
for truth.

@@

A sequence such as $\bar{h}{t}$ can be viewed both as a formula and as
a valuation of primitive propositions, which we use to describe world
state. The primitives are listed in fixed order, and left unmarked
(indicating true) or marked with the overbar (indicating false).
Since a coin is heads or tails but not both, we want to allow the
valuations $h\bar{t}$ and $\bar{h}t$, and disallow $ht$ and
$\bar{h}\bar{t}$.\footnote{Two Booleans constrained by a state formula, rather than one Boolean, are stipulated here to
 illustrate state formulas.
}  This is enforced by a {\em state formula}, which is
a Boolean formula, in this case the one given on the second line of
\ex{effect}.  Where {\sf{B}} is a set of state primitives and $\varphi$ is a
 constraint over {\sf{B}},
${\cal{A}}_{\mbox{\sf\tiny{B}}}^{\varphi}$ is the set of valuations of
{\sf{B}} that make formula $\varphi$ true. Valuations are called atoms,
because they correspond to the atoms of a Boolean algebra of tests
\citep{kozen2001automata}.

@@

Epik is a specification language for possible worlds models that
includes declarations of events and states, state formulas, effect
formulas, and additional information.  Figure \ref{epikfigure} shows
an Epik program that describes a possible worlds model for two agents
with information about one coin, events of the agents semi-privately
looking at the coin, and events of secretly turning the coin.  The
line beginning with {\tt state} enumerates ${\sf{B}}$. The line
beginning with {\tt restrict} gives the state formula.  The lines
beginning with {\tt event} declare events and their effect formulas.

Finally, the lines beginning with {\tt agent} define {\em event alternative}
relations for the epistemic agents in the model. Each clause with an arrow has a
single event symbol on the left, and a disjunction of alternative events on the
right of the arrow. The interpretation of Amy's alternatives for $\bpH$ (Bob
peeks at heads), is that when $\bpH$ happens, for Amy either $\bpH$ or $\bpT$ (Bob
peeks at tails) could be happening, indicating that she doesn't know whether Bob saw heads or tails, only that he peeked at the coin.

Similarly, Her alternatives for $\atTH$ (she turns the coin over from heads to
tails) are $\atTH$ and $\atHT$, indicating that she doesn't know, \emph{a
  priori}, whether she's turning the coin from $H$ to $T$ or from $T$ to $H$.
She has the same alternatives for the event $\atHT$. Conversely, When Bob
secretly turns the coin over, in event $\btTH$ or $\btHT$, she doesn't know
anything has happened, so her alternative is the ``no-operation'' or
``no-information'' event event $o$. Bob's event relation is symmetric.

@@

This paper focuses on defining a concrete possible worlds model from an Epik
specification. The models are an extension of guarded-string models for Kleene
Algebra with Tests (\textsc{Kat}). This algebraic theory has model classes including
guarded string models, relational models, finite models, and matrix models. Our
definitions and notation follow \cite{kozen2001automata}. We add syntax and
semantics to cover multi-agent epistemic semantics.

Guarded strings over a finite alphabet ${\sf{P}}$ are like ordinary strings, but
with atoms over a set ${\sf{B}}$ alternating with the symbols from ${\sf{P}}$.
In the algebra described by Figure \ref{epikfigure}, ${\sf{P}}$ is the set of
events $\{\apH,\apT,\bpH,\bpT,\atTH,\atHT,\btTH,\btHT,o\}$, and ${\sf{B}}$ is
$\{h,t\}$. As we already saw in \ex{effect}, $\mathcal{A}^\varphi_{\mathsf B}$
is $\{h\bar{t},\bar{h}t\}$, for which we use the shorthand $\{H,T\}$. A guarded
string over ${\sf{P}}$ and ${\sf{B}}$ is a string of events from ${\sf{P}}$,
alternating with atoms over ${\sf{B}}$, and beginning and ending with atoms. In
this construction, $w_1=H$, $w_2=Ha_1H$, $w_3=Ha_1Hb_1H$, and
$w_4=Ha_1Hb_1Ha_{10}Tb_{01}Hb_1H$.

@@

Rather than individual guarded strings, elements of a guarded string
model for \textsc{Kat} are sets of guarded strings.  In our application, these
elements have the interpretation of propositions, which are sets of possible
worlds.  In a free guarded string model for \textsc{Kat}, any event can be
adjacent to any atom in a guarded string that is an element of the
underlying set for the algebra.  We instead impose the constraints
coming from the state and effect formulas.  \ex{wf1} defines the
well-formed guarded strings determined by an Epik specification.  Condition (i)
says that each atom is consistent with the state constraint, and condition (ii)
says that each constituent token event $\alpha_ie_i\alpha_{i+1}$ is
consistent with the effect constraint on $e_i$.\footnote{
An alternative is to define equations such as
$\bar{\phi}=0$ (from the state formula $\phi$)
and $a_1\,{=}\,ha_{1}h$ (from the effect formula $h\mathbin{:}h$ for event $a_1$),
and construct a quotient algebra from the equivalence relation generated
by these equations.  This results in equating sets of guarded strings in the free algebra that
differ by guarded strings that are ill-formed according to the state and
effect formulas.  In the development in the text, we instead use
a set of guarded strings that are well-formed according to the state
and effect formulas as the representative of the equivalence class.
}

@@

The construction so far defines a set of worlds from an Epik specification.  Normally
the set is countably infinite, though some choices of effect formulas can result in a
finite set of worlds.  The next step is to define an alternative relation $R_a$ on
worlds for each agent $a$. This will result in a general modal frame
$\tuple{W,R_1,...,R_n,K}$ consisting of a set of worlds, a world-alternative relation
for each agent, and a set $K$ of propositions, where each proposition is a subset of
$W$ \citep{chagrov1997modal}.\footnote{As explained in the next section, $K$ will not
  be the power set of $W$, rather it will consists of the regular subsets of $W$.}

An Epik specification defines an alternative relation on bare events for each
agent $a$, which is notated ${R}_{a}$, and lifted to a relation $\hat{R}_a$ on
worlds. The basic idea is that when a world $w$ is incremented with an event
$e$, in the resulting world $w \diamond e$, epistemic alternatives for agent $a$
are of the form $w' \diamond e'$, where $w'$ is an alternative to for $a$ in
$w$, and $e'$ is and event-alternative to $e$ for $a$.\footnote{In this it is
  important that the event-alternative relation for an agent is constant across
  worlds. We anticipate that the definition given here produces results
  equivalent to what is found in literature on event alternatives in dynamic
  epistemic semantics, though we have not verified this. That literature
  primarily focuses on mapping an epistemic model for a single time and
  situation to another, and uses general first-order models, rather than guarded
  string models. See \citet{baltag1999logic}, \citet{van2007dynamic}, and
  articles in \citet{ditmarsch2015handbook}. This literature is motivated by
  epistemic logic and AI planning, rather than computable possible worlds models
  in natural language semantics. } This needs to be implemented in a way that
takes account of pre- and post-conditions for events. For this, our approach is
to refer the definition of well-formed guarded strings. \ex{raise1} defines an
epistemic alternative relation on worlds from an alternative relation on bare
events.

@@

Consider a scenario like the one from Figure 1, but with an additional agent
Cal. The base world $T b_0 T c_0 T$ is one where the coin is tails, and first
Bob looks at tails, and them Cal looks at tails. The first event $b_0$ has the
alternatives $b_0$ and $b_1$ for Amy, and the second event $c_0$ has the
alternatives $c_0$ and $c_1$ for Amy. This results in four combinations
$b_0c_0$, $b_0c_1$, $b_1c_0$, and $b_1c_1$. But some of these worlds are bogus,
such as $b_0c_1$, which says that Bob peeked at the coin and saw tails, and Cal
peeked at the same coin and impossibly saw heads. These contradictions are
filtered by post- and pre-conditions of events in the alternative world, so that
the set of alernatives for Amy in $T b_0 T c_0 T$ is $\{T b_0 T c_0 T,H b_1 H
c_1 H\}$, with two world-alternatives instead of four.

\begin{figure}
  \fbox{
    \begin{minipage}{\columnwidth}
      \(\begin{array}{rlrl}
      \multicolumn{4}{l}{\text{events}~~e\in \mathsf{P}} \\
      p,q &\multicolumn{3}{l}{::= e\,|\,\sigma\,|\,p+q\,|\,pq\,|\,p^*\,|\,\neg p\,|\, \Diamond_ap} \\
      \Box_a p & \triangleq \neg\Diamond_a\neg p &
      \bullet & \triangleq \sum_{e \in \mathsf P}e \\      
      p \wedge q &  \triangleq \neg(\neg p + \neg q) &
      p \to q & \triangleq \neg p + q
      \end{array} \)
  \end{minipage}}

@@

\section{The logical language of Epistemic \textsc{Kat}}
The standard language for Kleene algebra with tests has the signature
$\tuple{K,+,\cdot,*,\bar{},0,1}$ \citep{kozen2001automata}.  In a guarded string
model for \textsc{Kat}, $K$ is a set of sets of guarded strings, $+$ is set union, the
operation $\cdot$ is fusion product raised to sets, $*$ is Kleene star, the operation
$\;\bar{}\;$ is complement for tests, $0$ is the empty set, and $1$ is the set of
atoms.\footnote{$0$ has the dual role the identity for $+$ (union), and as False for
  operations on tests.  $1$ has the dual role of the identity for product (fusion
  product raised to sets), and True for tests.  } To this we add a unary
modal operation $\Diamond_a$ for each agent, and a unary complement operation $\neg$
on elements of $K$.  Intuitively, $\Diamond_a p$ is the set of
worlds where proposition $p$ is epistemically possible for agent $a$.  Propositional
complement is included because natural languages have sentence negation.  In
addition, universal box modalities are defined as duals of existential diamond
modalities.

With modalities and propositional negation added, the signature of $n$-agent
epistemic \textsc{Kat} is $\tuple{K,+,\cdot,*,\bar{},0,1,\neg,\Diamond_1,
  ..., \Diamond_n}$. Figure \ref{termfigure} defines the syntax of the language.
Juxtaposition is used for product.  Terms in this language
are used to represent the proposositional semantic values of English sentences.
\ex{interpretation1} gives some examples.  To explain the first one, $\bullet$ as
defined in Figure \ref{fig:epik-grammar} is the disjunction of the primitive events.
Since a world is a well-formed sequence of events, $\bullet^*$ is the set of worlds.
Multiplying by the state symbol $h$ in the term $\bullet^* h$ has the effect of
conjoining $h$ with the atom at the end of the world.  So $\bullet^* h$ is the set
of worlds where the coin ends heads-up.

@@

{\em Standard Epistemic Modalities.}
Using our existential modal primitive $\Diamond_a$, and the dual
encodings of $\Box_a$ and $\wedge$, we can encode the standard modal
operators expressing knowledge $(\K_a)$ and belief $(\B_a)$
as in \ex{encodings}.\footnote{Deeper analyisis of the lexical semantics
  of {\em know} requires adding modeling of presupposition
  \citep{collard2018finite}. The grammar fragment in Section 6 does not model
  the presupposition of {\em know}, except as
  an entailment.}

@@

Summing up, given an Epik specification $\epikspec$, term $p$ (as defined
syntactically in Figure \ref{termfigure}) is interpreted as a set of guarded strings
$\sem{p}^{\epikspec}$.  Let $K^{\epikspec}$ be the sets that are interpretations of
terms.  Then
$\tuple{K^{\epikspec},+,\cdot,*,\bar{},0,1,\neg,\Diamond_{a_1},...,\Diamond_{a_n}}$
is a concrete guarded string interpretation for the signature of epistemic \textsc{Kat}, with
operations as in Figure \ref{fig:interp2} (e.g. the binary operation $+$ is union,
and the unary operation $\Diamond_a$ is pre-image relative to $\hat{R}_a$).  This
provides a concrete $n$-agent general modal frame
$\tuple{\sem{\bullet^*}^{\epikspec}, \hat{R}_1,...,\hat{R}_n,K^{\epikspec}}$.  The
frame consists of a set of worlds, an epistemic-alternative relation for each agent,
and a set of propositions, each of which is a set of worlds.  It is used as a target
for natural-language interpretation in Section~\ref{sec:categorial-grammar}.

@@

\section{Translation into the finite state calculus}
The finite state calculus is an algebra of regular sets of strings and
regular relations between strings that was designed for use in
computationonal phonology and morphographemics
\citep{kaplan1994regular,beesley2003finite}.  Current implementations
allow for the definition of functions on regular sets and relations
\citep{hulden2009foma,linden2009hfst,karttunen2010update}.  Such definitions are used
here to construct of a model for epistemic \textsc{Kat} inside the finite state
calculus. We describe our translation from Epik logical terms to Fst terms
here.

@@

\begin{hangingpar}
\begin{verbatim}
define Cn(X,Y) 
  [[[X Y] & Wf0] .o. Squash].l;
\end{verbatim}
\textsc{Kat} product.
\end{hangingpar}

@@

\begin{hangingpar}
\begin{verbatim}
define Kst(X) St | Kpl(X);
\end{verbatim}
\textsc{Kat} Kleene plus and Kleene star. The Fst operation {\tt |} is union.
\end{hangingpar}
\end{tcolorbox}
\caption{\raggedright Definition in Fst of \textsc{Kat} product and star.
  %% Where {\tt X} and {\tt Y} are regular sets and
  %% {\tt R} and {\tt S} are regular relatiions,
  %% {\tt X{\&}Y} is the intersection of {\tt X} and {\tt Y},
  %% {\tt X|Y} is the union of {\tt X} and {\tt Y},
  %% {\tt \~{}X} is the complement of {\tt X},
  %% {\tt R{.o.}S} is the composition of {\tt R} and {\tt S},
  %% {\tt R{.}l} is the co-domain of {\tt R}, and
  %% {\tt {\$}X} is the set of strings that have a substring in $X$.   
}
\label{fstfigure1}
\end{figure}

@@

Translating the Epik terms $0$, $1$, $b$, and $e$ are
straightforward: we simply convert the atoms as previously described,
and decorate the events $e$ with their compatible atoms. For example $a_1$ becomes
an Fst term denoting
$\{01a_101\}$. Fst has built-in operations of union (\texttt{|}),
and intersection (\texttt{\&}), which are
the sum and intersection operations in the guarded string algebra as represented
in Fst.
Fst set difference (\texttt{-}) is used to define propositional complement as the
difference between the set of all worlds and the argument.

Defining \textsc{Kat} product using Fst's set-lifted string concatenation (denoted by
juxtaposition $\texttt{X}~\texttt{Y}$) requires more care. Naively concatenating
strings with atoms (Boolean vectors) at both ends doubles atoms at the juncture, and
does not enforce the requisite atom equality.  To implement \textsc{Kat} product, we define
the binary operation \texttt{Cn}, which concatenates strings in the string algebra,
removes strings with non-matching atoms, and then deletes the second of two atoms to
create a set of well-formed guarded strings.  See Figure \ref{fstfigure1}. {\tt Wf0}
is the set of ordinary strings that does not contain unequal pairs of atoms, as
defined using Fst's containment operator \texttt{\$}. The \texttt{Squash} relation
uses Fst's rewrite notation to delete atoms (elements of \texttt{St}) that are
preceded by another atom. \footnote{ This is a non-equal length regular relation. The
  finite state calculus includes such relations, and they can be used with relation
  composition and relation domain and co-domain.  They are restricted in that the
  complement and set difference for non-equal length relations is not defined.
  Epistemic alterative relations are equal-length relations.} This relation is
applied via the relational composition (\texttt{.o.}) and codomain (\texttt{.l})
operators.
%% Deletion is accomplished by a
%% re-write rule in the finite state calculus, which is a notation for
%% defining regular relations by contextually constrained substitutions.
%% In this case, is {\tt Squash} is a regular relation that deletes an
%% atom (a sequence of 0's and 1's of a certain length) when it follows
%% an atom.
